The unequal distribution of access to education and transport: insights from

### 2016-06-06 paper.Rmd
Fix typos

---
title: 'The unequal distribution of access to education and transport: insights from
  the São Paulo Metropolitan Region'
author: Ana I. Moreno-Monroy (Universitat Rovira i Virgili), Robin Lovelace (University of Leeds)  and Frederico R. Ramos (Fundação
  Getulio Vargas) 
output:
  pdf_document:
    keep_tex: yes
  html_document:
    keep_md: yes
  word_document: default
bibliography: school-access-bib.bib
---

@@

Inequalities in educational and transport infrastructure are mutually reinforcing: the right to mobility is intrinsically linked to the right to education. Travel to school options are vital for ensuring a more equitable supply of educational opportunity to diverse groups. Conversely, poor accessibility to deprived area can reinforce social inequalities, with long-term implications. Against this background, in this paper we propose a new way to masure school accessibility in local areas within an urban area, and aply it for the case of the São Paulo Metropolitan Region (SPMR).

@@

Our proposed competitive accessibility measure combines information about the place of residence of students, the spatial distribution of public schools, and public transport accessibility into a single measure. We start from a cumulative opportunity measure that counts the number of schools seats that can be reached within a 30 minute journey by public transport, and then build a competitive measure that takes into account the place of residency of teens. Travel-to-school times by public transport are based on actual commuting times obtained through the Google Travel Distance API. We calculate these measures for 633 areas within the SPMR. We use the competitive meausre to simulate the impact of a policy change in the location of public secondary schools, in order to, first, understand the extent of spatial inequalities in school accessibility by public transport, and second, estimate the effect of a policy aiming at concentrating public secondary education provision on school accessibility. We find that closing down schools in areas with lower than average provision is highly regressive: it negatively impacts students with the lowest accessibility levels, accentuating existing inequalities.

@@

Figure XX shows the spatial distribution of the competitive accessibility measure. Although it largely follows the distribution of the cumulative index, it corrects the value in some areas based on the place of residence of students across the SPRM. To interpret the value of the index, recall that if all students in an area attended secondary schools in public secondary schools in their area (i.e., no student would travel outside her own area to attend school), the index would take the value of 1. Then, in areas with a value lower than one, the provision of public schools and public transport is such that there is local underprovision, in the sense that the number of schools seats that can be reached from the area is smaller than the number of teens that can potentially reach the area (the opposite is true for values higher than one). Note that some relatively close by areas, especially in the eastern part of the city, have striking differences in the level of accessibility. In some cases, an area can have twice the level of accessibility than the most proximate area. This can be due to differences in the local provision of public schools, public transport or the number of students living within the reach of the areas.

@@

To understand the effects of changes in public school provision on school accessiblity, we conduct a comparative static analysis using our competitive index. To simulate the effects of a policy that seeks to spatially concentrate students in larger public schools by closing down smaller schools, we assign students in areas with lower than average enrolment to the next area with higher than average enrolment, while keeping the distribution of teens and public transport constant. We then compare the changes in school accessibility for different quartiles before and after this change. Table XX summarizes the results.

@@

The policy reduces further the accessibility level of areas in the first and second quantiles of accessibility. Areas in the first quantile experience an average reduction of 21% in school accessibility levels as a result of the changes in the spatial distribution of schools. In the mean time, areas that had high levels of accesibility to start with, experience an increase. The results suggest that concentrating the provision of public secondary schools accentuates the already large differences in accesibility levels across areas.

@@

The policy experiment results suggest that a policy that concentrates the provision of public schools spatially is regressive, in the sense that teens living in areas that had the lowest levels of accessibility are the ones who experience the largest negative impact from the policy. They are likely to be worse off compared to an already difficult situation of low local schooling provision, low access by public transport and/or high competition by teens in other areas. In practice, such a change would probabily imply larger commuting times for students residing in areas with low accessibility, which can have further consequences in terms of attendance, drop-out likelihood and performance.

@@

Our study can be further extended to account for other modes of school commute, such as active travel. It can also benefit from a finer spatial dissagregation to capture within area inequalities, which may be substantial and are currently hidden away in area aggregates. A future extension of the index can incorporate a measure of school quality, which in so far assumed away. It will also be highly informative to calculate the proposed indices for other contexts and for different time periods in order to understand the true magnitude of the extent of inequality across individuals living in different areas of the SPMR.

We conclude that extending the concept of local accessibility indicators to education can help to both contest and constructively tackle embedded social inequalities. Contestation can be seen as a powerful tool to prevent the implementation of regressive policies that threaten to accentuate already large disparities. Quantitative measures such as the ones proposed here help summarizing multiple dimensions of inequality into a single, intterpretable measure. Having this tool may be useful for both policy makers and social groups in discussing and negotiating policies with a highly sensitive social component and evident short and long-term consequences for those affected.

>>>>

---
title: 'The unequal distribution of access to education and transport: insights from
  the São Paulo Metropolitan Region'
author: Ana I. Moreno-Monroy (Universitat Rovira i Virgili), Robin Lovelace (University
  of Leeds)  and Frederico R. Ramos (Fundação Getulio Vargas)
output:
  word_document: default
  pdf_document:
    keep_tex: yes
  html_document:
    keep_md: yes
bibliography: school-access-bib.bib
---

@@

Inequalities in educational and transport infrastructure are mutually reinforcing: the right to mobility is intrinsically linked to the right to education. Travel to school options are vital for ensuring a more equitable supply of educational opportunity to diverse groups. Conversely, poor accessibility to deprived area can reinforce social inequalities, with long-term implications. Against this background, in this paper we propose a new way to measure school accessibility in local areas within an urban area, and ply it for the case of the São Paulo Metropolitan Region (SPMR).

@@

Our proposed competitive accessibility measure combines information about the place of residence of students, the spatial distribution of public schools, and public transport accessibility into a single measure. We start from a cumulative opportunity measure that counts the number of schools seats that can be reached within a 30 minute journey by public transport, and then build a competitive measure that takes into account the place of residency of teens. Travel-to-school times by public transport are based on actual commuting times obtained through the Google Travel Distance API. We calculate these measures for 633 areas within the SPMR. We use the competitive measure to simulate the impact of a policy change in the location of public secondary schools, in order to, first, understand the extent of spatial inequalities in school accessibility by public transport, and second, estimate the effect of a policy aiming at concentrating public secondary education provision on school accessibility. We find that closing down schools in areas with lower than average provision is highly regressive: it negatively impacts students with the lowest accessibility levels, accentuating existing inequalities.

@@

Figure XX shows the spatial distribution of the competitive accessibility measure. Although it largely follows the distribution of the cumulative index, it corrects the value in some areas based on the place of residence of students across the SPRM. To interpret the value of the index, recall that if all students in an area attended secondary schools in public secondary schools in their area (i.e., no student would travel outside her own area to attend school), the index would take the value of 1. Then, in areas with a value lower than one, the provision of public schools and public transport is such that there is local under-provision, in the sense that the number of schools seats that can be reached from the area is smaller than the number of teens that can potentially reach the area (the opposite is true for values higher than one). Note that some relatively close by areas, especially in the eastern part of the city, have striking differences in the level of accessibility. In some cases, an area can have twice the level of accessibility than the most proximate area. This can be due to differences in the local provision of public schools, public transport or the number of students living within the reach of the areas.

@@

To understand the effects of changes in public school provision on school accessibility, we conduct a comparative static analysis using our competitive index. To simulate the effects of a policy that seeks to spatially concentrate students in larger public schools by closing down smaller schools, we assign students in areas with lower than average enrolment to the next area with higher than average enrolment, while keeping the distribution of teens and public transport constant. We then compare the changes in school accessibility for different quartiles before and after this change. Table XX summarizes the results.

@@

The policy reduces further the accessibility level of areas in the first and second quantiles of accessibility. Areas in the first quantile experience an average reduction of 21% in school accessibility levels as a result of the changes in the spatial distribution of schools. In the mean time, areas that had high levels of accessibility to start with, experience an increase. The results suggest that concentrating the provision of public secondary schools accentuates the already large differences in accessibility levels across areas.

@@

The policy experiment results suggest that a policy that concentrates the provision of public schools spatially is regressive, in the sense that teens living in areas that had the lowest levels of accessibility are the ones who experience the largest negative impact from the policy. They are likely to be worse off compared to an already difficult situation of low local schooling provision, low access by public transport and/or high competition by teens in other areas. In practice, such a change would probably imply larger commuting times for students residing in areas with low accessibility, which can have further consequences in terms of attendance, drop-out likelihood and performance.

@@

Our study can be further extended to account for other modes of school commute, such as active travel. It can also benefit from a finer spatial disaggregation to capture within area inequalities, which may be substantial and are currently hidden away in area aggregates. A future extension of the index can incorporate a measure of school quality, which in so far assumed away. It will also be highly informative to calculate the proposed indices for other contexts and for different time periods in order to understand the true magnitude of the extent of inequality across individuals living in different areas of the SPMR.

We conclude that extending the concept of local accessibility indicators to education can help to both contest and constructively tackle embedded social inequalities. Contestation can be seen as a powerful tool to prevent the implementation of regressive policies that threaten to accentuate already large disparities. Quantitative measures such as the ones proposed here help summarizing multiple dimensions of inequality into a single, interpretable measure. Having this tool may be useful for both policy makers and social groups in discussing and negotiating policies with a highly sensitive social component and evident short and long-term consequences for those affected.
